---
title: "Lab 9 - HPC"
author: "Rayce Ramsay"
output: github_document
link-citations: yes
---

# Learning goals

In this lab, you are expected to practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r eval=FALSE, echo=FALSE}
# install any missing packages
# install.packages("microbenchmark")
```

```{r}
library(parallel)
library(microbenchmark)
library(tidyverse)
```


## Problem 1

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

1. Markov Chain Monte Carlo: `rstan` can be used for Bayesian modelling, while `RcppParallel` can be used for parallel mcmc sampling

2. Image Denoising: `mclapply` and `parLapply` can be used for parallelized loop ops

3. Cross-Validation in machine learning: supported by `doParallel`

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(),
  fun1alt()
)
```

How much faster?

The alternative implementation is roughly 7 times faster on average.


2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
# Inefficient because it uses apply() which loops through each column giving a slower function.
fun2 <- function(x) {
  apply(x, 2, max)
}

# Avoids loops and directly extracts max value
fun2alt <- function(x) {
  x[cbind(max.col(t(x)), 1:ncol(x))]
}

# Benchmarking
bench = microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x),
  unit = "us"
)
bench
```

_Answer here with a plot._

```{r}
plot(bench)

autoplot(bench) + 
  theme_minimal()
```


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
    # ncpus specify that we are using multiple CPU cores
    # PSOCK for parallel socket cluster
  cl = makePSOCKcluster(ncpus)
  # on.exit(stopCluster(cl)) # makes the clusters automatically shutdown to prevent memory leak
  
  # STEP 2: GOES HERE
  clusterExport(cl, varlist = c("dat", "stat", "idx"), envir = environment())
  
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  # ans <- lapply(seq_len(R), function(i) {
  #   stat(dat[idx[,i], , drop=FALSE])
  # })
  ans = parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[, i], , drop = FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl)
  
  ans
  
}
```

1. Use the previous pseudocode, and make it work with `parallel`. Here is just an example for you to try:

```{r p3-test-boot}
# Bootstrap of a linear regression model
# my_stat <- function_for_lm 
my_stat = function(d) coef(lm(y~x, data = d))

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)

# Check if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R=R, ncpus = 4)
qs <- c(.025, .975)

cat("Ordinary Least Squares: \n")
print(ans0)
cat("\n")
cat("Bootstrap CI: \n")
print(t(apply(ans1, 2, quantile, probs = qs)))
```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3}
parallel::detectCores()
system.time(my_boot(dat = data.frame(x, y), my_stat, R= 4000, ncpus = 1L))
system.time(my_boot(dat = data.frame(x, y), my_stat, R= 4000, ncpus = 8L))
```

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("[full-path-to-your-Rmd-file.Rmd]")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


