---
title: "Lab 08 - Text Mining/NLP"
author: "Rayce Ramsay"
output: github_document
always_allow_html: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = T, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r, warning=FALSE}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)
library(reshape2)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r}
mt_samples %>%
  count(medical_specialty, sort = TRUE) %>%
  ggplot(aes(x = reorder(medical_specialty, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Count of Medical Specialties",
    x = "Medical Specialty",
    y = "Number of Transcriptions"
  ) +
  theme_bw()
```
The categories don't seem to be related too much. They are not evenly distributed (surgery has over 1000 while immunology appears to have 0).

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

```{r}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  group_by(word) |>
  summarize(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)

tokens |>
  ggplot(aes(fct_reorder(word, word_frequency), word_frequency)) +
  geom_bar(stat = 'identity', fill = "dodgerblue") +
  coord_flip() +
  labs(
    title = "Top 20 Most Frequent Words",
    x = "Word",
    y = "Frequency"
  ) +
  theme_bw()
```

```{r}
tokens |>
  count(word, sort=TRUE) |>
  wordcloud2(size = 0.5, color = "random-light", backgroundColor = "dodgerblue")
```


We can see that the most frequent words are stopwords. This is expected as stopwords by nature are supposed to be the most common. It is hard to extract much insight from this.

---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?

```{r}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)

custom_stopwords <- c("mm", "mg", "also", "using")

tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  filter(!word %in% stopwords("en"),
         !word %in% custom_stopwords,
         !str_detect(word, "^[0-9]+$")) |> 
  count(word, sort = TRUE) |>
  slice_head(n = 20)

tokens |>
  ggplot(aes(x = fct_reorder(word, n), y = n)) +
  geom_col(fill = "dodgerblue") +
  coord_flip() +
  labs(
    title = "Top 20 Words (Without Stopwords & Numbers)",
    x = "Word",
    y = "Frequency"
  ) +
  theme_bw()
```

```{r}
tokens |>
  count(word, sort=TRUE) |>
  wordcloud2(size = 0.4, color = "random-light", backgroundColor = "dodgerblue")
```

After removing stopwords, we can see that there are a lot of medical terms in the transcriptions. This suggests that the text probably has to do with medical related topics.

---

## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r}
stopwords2 <- c(stopwords("en"), "mm", "mg", "also", "using")

sw_start <- paste0("^", paste(stopwords2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stopwords2, collapse="$| "), "$")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!str_detect(ngram, "^[0-9]+$"))|>
  filter(!str_detect(ngram, sw_start))|>
  filter(!str_detect(ngram, sw_end))|>
  count(ngram, sort = TRUE)
```

```{r}
tokens_bigram20 <- tokens_bigram |> head(20)

tokens_bigram20 |>
  ggplot(aes(x = fct_reorder(ngram, n), y = n)) +
  geom_col(fill = "dodgerblue") +
  coord_flip() +
  labs(
    title = "Top 20 Most Frequent Bi-Grams",
    x = "Bi-Gram",
    y = "Frequency"
  ) +
  theme_bw()
```

```{r}
tokens_trigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 3) |>
  filter(!str_detect(ngram, "^[0-9]+$"))|>
  filter(!str_detect(ngram, sw_start))|>
  filter(!str_detect(ngram, sw_end))|>
  count(ngram, sort = TRUE)
```

```{r}
tokens_trigram20 <- tokens_trigram |> head(20)

tokens_trigram20 |>
  ggplot(aes(x = fct_reorder(ngram, n), y = n)) +
  geom_col(fill = "dodgerblue") +
  coord_flip() +
  labs(
    title = "Top 20 Most Frequent Tri-Grams",
    x = "Tri-Gram",
    y = "Frequency"
  ) +
  theme_bw()
```

We seem to get phrases that can be understood out of context better when we look at tri-grams versus bi-grams.

---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r}
library(stringr)

word_of_interest <- "patient"

tokens_bigram_filtered <- tokens_bigram |>
  filter(str_detect(ngram, paste0("\\b", word_of_interest, "\\b"))) |>
  mutate(
    word = str_remove(ngram, paste0("\\b", word_of_interest, "\\b")), 
    word = str_remove_all(word, "\\s")
  ) |>
  count(word, sort = TRUE) |>
  head(20)

tokens_bigram_filtered$word <- as.character(tokens_bigram_filtered$word)

tokens_bigram_filtered |>
  ggplot(aes(x = fct_reorder(word, n), y = n)) +
  geom_col(fill = "dodgerblue") +
  coord_flip() +
  labs(
    title = paste0("Top 20 Words Appearing Before or After '", word_of_interest, "'"),
    x = paste0("Word appearing with '", word_of_interest, "'"),
    y = "Frequency"
  ) +
  theme_bw()

```

---


## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?


```{r}
top_words_by_specialty <- mt_samples |>
  unnest_tokens(word, transcription) |>
  filter(
    !word %in% stopwords2, 
    !str_detect(word, "^[0-9]+$")
  ) |>
  count(medical_specialty, word, sort = TRUE) |>
  group_by(medical_specialty) |>
  slice_max(n, n = 5) |>  
  ungroup() |>
  arrange(medical_specialty, -n)

top_words_by_specialty
```

We can view the most common words in each speciality using the table above. For example, the most commonly used word in the "Allergy" category is "allergies" while the most common word in the "Bariatrics" category is patient.

```{r}
top_words <- mt_samples |>
  unnest_tokens(word, transcription) |>
  filter(
    !word %in% stopwords2, 
    !str_detect(word, "^[0-9]+$")
  ) |>
  count(word, sort = TRUE) |>
  arrange(-n) |>
  head(5)
top_words
```

The top 5 most used words (regardless of category) are "patient", "right", "left", "procedure", and "placed."

## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r}
transcripts_dtm <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by="word") |>
  filter(!str_detect(word, "^[0-9]+$"),
         !word %in% c("mm", "mg", "also", "using")) |>
  DocumentTermMatrix()
    
transcripts_dtm <- as.matrix(transcripts_dtm) 

transcripts_lda <- LDA(transcripts_dtm, k=5, control = list(seed=1234))
transcripts_lda

transcripts_tope_terms <- 
  tidy(transcripts_lda, matrix = "beta") |>
  filter(!str_detect(term, "^[0-9]+$")) |>
  group_by(topic) |>
  slice_max(beta, n=10) |>
  ungroup() |>
  arrange(topic, -beta)

transcripts_tope_terms |> 
  mutate(term=reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill=factor(topic)))+
  geom_col(show.legend=FALSE)+
  facet_wrap(~topic, scales="free")+
  scale_y_reordered()+
  theme_bw()
```